### 2021-08-10 manuscript/ex_article.tex
.

% REQUIRED
\begin{abstract}
\red{
This paper concerns the modeling and numerical simulation of the process of speciation. In particular, given conditions for which one or more speciation events within an ecosystem occur, our aim is to develop the necessary modeling and simulation tools. Care is also taken to establish a solid mathematical foundation on which our modeling framework is built. This is the subject of the first half of the paper. The second half is devoted to developing a multi-scale framework for eco-evolutionary modeling, where the relevant scales are that of species and individual/population, respectively. The species level model we employ can be considered as an extension of the classical Lotka-Volterra model, where in addition to the species abundance, the model also governs the evolution of the species mean traits and species trait covariances, and in this sense generalizes the purely ecological Lotka-Volterra model to an eco-evolutionary model. Although the model thus allows for evolving species, it does not (by construction) allow for the branching of species, i.e., speciation events. The reason for this is related to that of separate scales; the unit of species is too coarse to capture the fine-scale dynamics of a speciation event. Instead, the branching species should be regarded as a population of individuals moving along a selection of trait axes (i.e., trait-space). For this, we employ a trait-specific population density model governing the dynamics of the population density as a function of evolutionary traits. At this scale there is no a priori definition of species, but both species and speciation may be defined a posteriori as e.g., local maxima and saddle points of the population density, respectively.  
%As long as the identification of distinct species remains non-ambiguous, both the fine-scale (population) and coarse-scale (species) level models describe the same system dynamics (i.e., evolution of the same ecosystem), while  
%In this sense, we regard both species and speciation as emergent phenomena, ultimately due to individuals interacting within a changing environment. 
Hence, a system of interacting species can be described at the species level, while for branching species a population level description is necessary. 
%Assuming the existence of statistical moments of the population density function, and that the individuals of each species follows a known statistical distribution,  
The coupling of the species and population scales into a multi-scale framework is the subject of the second part of the paper. Moreover, since the population level model is formulated as a PDE, we first establish the well-posedness in the time-discrete setting, and then derive the a posteriori error estimates which provides a fully computable upper bound on an energy-type error, including also for the case of general smooth distributions. Our coupling strategy is twofold; monitoring the species in the system and detecting speciation in advance (using the a posteriori error bounds), and, splitting the diverging `parent'-species into new `child'-species. 
%Since the problem at hand is related to that of separate scales, it is convenient to have access to a fine-scale description of the same eco-evolutionary system. In practice, our strategy amounts to coupling the species and population level models in a multi-scale framework. This includes well-posedness and a posteriori error estimates of the time discrete population density model. 
Several numerical tests validate our framework in practice.
}
\end{abstract}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
\label{sec:intro}
Mathematical models have a long history in ecology and evolutionary biology, with the most famous example being the Lotka-Volterra model~\cite{leigh1968ecological}, which describes the interaction of two species (commonly referred to as predator and prey) at a timescale where individual traits remain constant, i.e., no evolution occuring. Going beyond strictly ecological models, the interaction between ecology and evolution has been studied extensively in the so-called ``adaptive dynamics" literature (see e.g.,~\cite{dieckmann1997can,dieckmann2007adaptive,gavrilets2004fitness,geritz2000adaptive}, and the references therein). In particular, the concept known as ``adaptive speciation", i.e., the idea that a series of small adaptive changes in the traits of individuals over long enough time leads to a diversity of species, has received considerable attention~\cite{dieckmann1999origin,dieckmann2004adaptive,doebeli1996quantitative,doebeli2000evolutionary,doebeli2003speciation,rosenzweig1978competitive}. In the current work, we are studying the process of adaptive speciation at the level of model formulation where the \emph{species} is regarded as the fundamental unit of the eco-evolutionary system, and where the species dynamics is emergent from the dynamics at the population level.

@@

The species-interaction model we take as the starting point is formulated as a system of ordinary differential equations (ODE) describing the dynamics of a biological system in which any finite number of species interact and evolve adaptively over evolutionary time (referred to in the sequel as the \emph{species level}, SLM, or \emph{macro-scale} model). This model extends the Lotka-Volterra system by representing each species not only by its abundance, but also by its mean traits coordinate and trait covariance matrix. It originates in a recent work by two of the authors~\cite{nordbotten2016asymmetric}, wherein the model equations were derived by applying a moment closure averaging technique on a deterministic trait-specific population density model (referred to in the sequel as the \emph{population level}, PLM, or \emph{micro-scale} model), \red{under the assumptions that (A1) distinct species can be identified at all times, and (A2) that the species distributions are of a known statistical quality (in this case, the normal distribution.} see also related derivations in~\cite{debarre2014multidimensional,sasaki2011oligomorphic}). The population level model is formulated as a partial differential equation (PDE), and governs the abundance density as a function of evolutionary traits. Moreover, the authors showed that the species and population level models are consistent, i.e., they describe the same system dynamics as long as the system behavior is such that both models are valid. Still, these descriptions are fundamentally different in that they operate on separate scales; the species level model operates with the coarser unit of distinct species, while the population level model operates with the finer unit of the individual (i.e., the category of species is not imposed upon the individuals who make up the population). Thus, the species level model described above approximates the system dynamics governed by the population level model, under the conditions that the population can be grouped into distinct species, \red{and that each species' distribution conforms to a normal distribution}. Note also that due to the deterministic nature of the population level model, this property is also inherited at the species level.

@@

Based on our results, we find that the multi-scale approach gives the better approximation of the reference PLM solution post-speciation, and thus conclude that a multi-scale approach is indeed necessary for robust eco-evolutionary modeling at the level of species interaction, capable of handling speciation events. Although we base our developments on the specific models described above, our methodology is more general; it is applicable on any species interaction model which, including the species abundance, also incorporates the evolution of species mean traits and species trait covariance in some way (thus allowing for the reconstruction of population density distributions), whether the model is based on deterministic or stochastic dynamics. The figure below shows a schematic representation of the micro/macro-scale coupling strategy for speciation events.

>>>>

% REQUIRED
\begin{abstract}
\red{
This paper concerns the modeling and numerical simulation of the process of speciation. In particular, given conditions for which one or more speciation events within an ecosystem occur, our aim is to develop the necessary modeling and simulation tools. Care is also taken to establish a solid mathematical foundation on which our modeling framework is built. This is the subject of the first half of the paper. The second half is devoted to developing a multi-scale framework for eco-evolutionary modeling, where the relevant scales are that of species and individual/population, respectively. The species level model we employ can be considered as an extension of the classical Lotka-Volterra model, where in addition to the species abundance, the model also governs the evolution of the species mean traits and species trait covariances, and in this sense generalizes the purely ecological Lotka-Volterra model to an eco-evolutionary model. Although the model thus allows for evolving species, it does not (by construction) allow for the branching of species, i.e., speciation events. The reason for this is related to that of separate scales; the unit of species is too coarse to capture the fine-scale dynamics of a speciation event. Instead, the branching species should be regarded as a population of individuals moving along a selection of trait axes (i.e., trait-space). For this, we employ a trait-specific population density model governing the dynamics of the population density as a function of evolutionary traits. At this scale there is no a priori definition of species, but both species and speciation may be defined a posteriori as e.g., local maxima and saddle points of the population density, respectively.  
%As long as the identification of distinct species remains non-ambiguous, both the fine-scale (population) and coarse-scale (species) level models describe the same system dynamics (i.e., evolution of the same ecosystem), while  
%In this sense, we regard both species and speciation as emergent phenomena, ultimately due to individuals interacting within a changing environment. 
Hence, a system of interacting species can be described at the species level, while for branching species a population level description is necessary. 
%Assuming the existence of statistical moments of the population density function, and that the individuals of each species follows a known statistical distribution,  
Our multi-scale framework thus consists of coupling the species and population level models where speciation events are detected in advance and then resolved at the population scale until the branchin is complete.
%The coupling of the species and population scales into a multi-scale framework is the subject of the second part of the paper. 
Moreover, since the population level model is formulated as a PDE, we first establish the well-posedness in the time-discrete setting, and then derive the a posteriori error estimates which provides a fully computable upper bound on an energy-type error, including also for the case of general smooth distributions (which will be useful for the detection of speciation events). 
%Our coupling strategy is twofold; monitoring the species in the system and detecting speciation in advance (using the a posteriori error bounds), and, splitting the diverging `parent'-species into new `child'-species. 
%Since the problem at hand is related to that of separate scales, it is convenient to have access to a fine-scale description of the same eco-evolutionary system. In practice, our strategy amounts to coupling the species and population level models in a multi-scale framework. This includes well-posedness and a posteriori error estimates of the time discrete population density model. 
Several numerical tests validate our framework in practice.
}
\end{abstract}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
\label{sec:intro}
Mathematical models have a long history in ecology and evolutionary biology, with the most famous example being the Lotka-Volterra model~\cite{leigh1968ecological}, which describes the interaction of two species (commonly referred to as predator and prey) at a timescale where individual traits remain constant, i.e., no evolution. Going beyond strictly ecological models, the interaction between ecology and evolution has been studied extensively in the so-called ``adaptive dynamics" literature (see e.g.,~\cite{dieckmann1997can,dieckmann2007adaptive,gavrilets2004fitness,geritz2000adaptive}, and the references therein). In particular, the concept known as ``adaptive speciation", i.e., the idea that a series of small adaptive changes in the traits of individuals over long enough time leads to a diversity of species, has received considerable attention~\cite{dieckmann1999origin,dieckmann2004adaptive,doebeli1996quantitative,doebeli2000evolutionary,doebeli2003speciation,rosenzweig1978competitive}. In the current work, we are studying the process of adaptive speciation at the level of model formulation where the \emph{species} is regarded as the fundamental unit of the eco-evolutionary system, and where the species dynamics is emergent from the dynamics at the population\red{/individual} level.

@@

The species-interaction model we take as the starting point is formulated as a system of ordinary differential equations (ODE) describing the dynamics of a biological system in which any finite number of species interact and evolve adaptively over evolutionary time (referred to in the sequel as the \emph{species level}, SLM, or \emph{macro-scale} model). This model extends the Lotka-Volterra system by representing each species not only by its abundance, but also by its mean traits coordinate and trait covariance matrix. It originates in a recent work by two of the authors~\cite{nordbotten2016asymmetric}, wherein the model equations were derived by applying a moment closure averaging technique on a deterministic trait-specific population density model (referred to in the sequel as the \emph{population level}, PLM, or \emph{micro-scale} model), \red{under the assumptions that (A1) distinct species can be identified at all times, and (A2) that the species distributions are of a known statistical quality (in this case, the normal distribution)}. See also related derivations in e.g.,~\cite{debarre2014multidimensional,sasaki2011oligomorphic}. The population level model is formulated as a partial differential equation (PDE), and governs the abundance density as a function of evolutionary traits. Moreover, the authors showed in~\cite{nordbotten2016asymmetric} that the species and population level models are consistent, i.e., they describe the same system dynamics as long as the system behavior is such that both models are valid. Still, these descriptions are fundamentally different in that they operate on separate scales; the species level model operates with the coarser unit of distinct species, while the population level model operates with the finer unit of the individual (i.e., the category of species is not imposed upon the individuals who make up the population). Thus, the species level model described above approximates the system dynamics governed by the population level model, under the conditions that the population can be grouped into distinct species, \red{and that each species' distribution conforms to a normal distribution}. Note also that due to the deterministic nature of the population level model, this property is also inherited at the species level.

@@

Based on our results, we find that the multi-scale approach gives the better approximation of the reference PLM solution post-speciation, and thus conclude that a multi-scale approach is indeed necessary for robust eco-evolutionary modeling at the level of species interaction, capable of handling speciation events. Although we base our developments on the specific models described above, our methodology is more general; it is applicable on any species interaction model which, including the species abundance, also incorporates the evolution of species mean traits and species trait covariance in some way (thus allowing for the reconstruction of population density distributions), whether the model is based on deterministic or stochastic dynamics. Figure \ref{fig:jan} below shows a schematic representation of the micro/macro-scale coupling strategy for speciation events.
